---
title: "Have Pokemon gotten stronger over the generations? (R)"
author: "Vaibhav Korrapati"
date: 2026-09-28
categories: [r, pokemon, data analysis]
---

In this post, we're going to analyze whether there's been a "power creep" in
Pokemon across generations. We're going to do that by calculating base stat totals (BSTs)
and do a comparison across generations.

## Data

The data is from [PokeAPI](https://pokeapi.co/)'s open dataset, taken from
this repository: [GitHub](https://github.com/PokeAPI/pokeapi/tree/master/data/v2/csv)

We use two tables: `pokemon_species.csv` (one row per species, with legendary
and mythical flags) and `pokemon_stats.csv` (one row per Pokemon per stat).

The same tables are used in my [Python post](../pokemon-legendary-python/index.qmd) analysing cumulative stats.

```{r}
#| label: load-data
#| message: false
#| warning: false
library(readr)
library(dplyr)
library(ggplot2)

species <- read_csv("../../data/pokeapi/pokemon_species.csv")
pokemon_stats <- read_csv("../../data/pokeapi/pokemon_stats.csv")

head(pokemon_stats)
```

## Calculating base stat totals

Summing `base_stat` for each `pokemon_id` gives its BST, which we then join with the
species table to get an appropriate mapping. Legendary and mythical Pokemon are removed, because the number
of legendaries varies a lot between generations and would skew the averages and we saw in
the previous post that they have a much higher BST on average. The main intent here is to compare
analyze power creep for regular Pokemon.

```{r}
#| label: compute-bst
bst <- pokemon_stats |>
  group_by(pokemon_id) |>
  summarize(bst = sum(base_stat))

regular <- species |>
  left_join(bst, by = c("id" = "pokemon_id")) |>
  filter(is_legendary == 0, is_mythical == 0) |>
  select(identifier, generation_id, bst)

head(regular)
```

## BST by generation

```{r}
#| label: tbl-bst-by-generation
#| tbl-cap: "Base stat total of regular (non-legendary, non-mythical) Pokemon by generation."
by_generation <- regular |>
  group_by(generation_id) |>
  summarize(
    n = n(),
    mean_bst = round(mean(bst), 1),
    median_bst = median(bst)
  )

knitr::kable(by_generation)
```

```{r}
#| label: fig-bst-by-generation
#| fig-cap: "Base stat totals of regular Pokemon by generation. Each box shows the middle 50% of Pokemon; the line inside is the median."
#| fig-alt: "Boxplots of base stat total for generations 1 to 9. Medians sit between about 405 and 450 for generations 1 to 6 and rise to between 470 and 483 for generations 7 to 9."
ggplot(regular, aes(x = factor(generation_id), y = bst)) +
  geom_boxplot(fill = "lightblue") +
  labs(x = "Generation", y = "Base stat total")
```

## Takeaways

There's a power creep, but its fairly even until Gen 3. @tbl-bst-by-generation
shows the average regular Pokemon actually got slightly weaker in Gen 2
and 3, then rose unevenly after that, with Gen 9 having the strongest. The
median in @fig-bst-by-generation shows a bigger jump, from around 405 in
Generation 1 to over 480 in Generation 9. The boxes still overlap heavily though so
there are plenty of strong old gen Pokemon and weak new gen ones.
